---
title: "Zena Creek Steelhead Habitat Potential"
author: Mike Ackerman
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::html_document2:
    theme: yeti
---

```{r setup, echo = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  #fig.path = "../output/figures/",
  #dev = c("pdf", "png"),
  dpi = 600
)

# replace NA w/ "-" in tables
options(knitr.kable.NA = "-")

```

```{r load-libraries}
# for formatting
library(kableExtra)

# for analysis
library(tidyverse)
library(sf)
library(ggspatial)
library(here)
library(janitor)
library(scales)

```

```{r load-data}
# load qrf datasets
load(here("output/prepped_snake_juv_sum_qrf.rda")) # prepped juvenile summer rearing (untrimmed to chnk & sthd extents)
qrf_redd_sf = get(load(here("output/prepped_snake_redd_qrf_untrimmed.rda"))) ; rm(qrf_sf) # prepped redd dataset (untrimmed to chnk & sthd extents)

```

```{r stream-segments}
# zena stream segments below impoundment dam
blw_ss = c(1083665, 1084335, 1097061, 1126937)

# zena stream segments opened to impoundment dam
abv_ss = c(997571, 1024185, 1039026, 1043936, 1048022, 1051588, 1051904, 1058839, 1060105, 1068759, 1070992, 1074093, 1076449, 1079437,
           1079546, 1093549, 1095914, 1101227, 1102813, 1103590, 1103840, 1104014, 1109010, 1110163, 1113965, 1114957, 1115362, 1118244, 1119930,
           1123030, 1124842, 1124897, 1129214, 1016572, 1069818, 1078288)

```

```{r analysis}
# trim qrf datasets to stream segments in blw_ss & abv_ss
zena_qrf_sf = qrf_redd_sf %>%
  mutate(model = "redd") %>%
  bind_rows(qrf_juv_sum_sf %>%
              mutate(model = "juv_sum")) %>%
  filter(unique_id %in% c(blw_ss, abv_ss)) %>%
  mutate(status = case_when(
    unique_id %in% blw_ss ~ "current",
    unique_id %in% abv_ss ~ "potential",
    TRUE ~ NA_character_
  ))

# summarize capacity below impoundment and potential new habitat
zena_df = zena_qrf_sf %>%
  st_drop_geometry() %>%
  group_by(model, status) %>%
  summarize(length_m = sum(reach_leng_m),
            capacity = sum(sthd_per_m * reach_leng_m),
            capacity_per_km = capacity / (length_m / 1000),
            .groups = "drop")

```

Currently, available spawning habitat for steelhead appears to be a limiting factor for production in Zena Creek. Adults are limited to spawning within the lower approximately `r zena_df$length_m[zena_df$status == "current" & zena_df$model == "redd"]` meters below the impoundment which is estimated to support only `r round(zena_df$capacity[zena_df$status == "current" & zena_df$model == "redd"], 0)` steelhead redds (Table \@ref(tab:table)). Removing the current barrier would provide an additional `r round(zena_df$length_m[zena_df$status == "potential" & zena_df$model == "redd"],0)` meters of accessible spawning habitat supporting a potential of `r round(zena_df$capacity[zena_df$status == "potential" & zena_df$model == "redd"], 0)` additional redds (Table \@ref(tab:table)). Further, we estimate that the additional accessible habitat would provide a greater than 6-fold increase in available rearing habitat for steelhead fry and parr. Our evaluation of habitat capacity in Zena Creek uses results from the quantile random forest models described by See et al. (2021).


```{r table}
zena_tbl = zena_df %>%
  clean_names("title") %>%
  select(Status,
         Model,
         `Stream Length (m)` = `Length m`,
         Capacity) %>%
  mutate(across(where(is.numeric),
                ~ comma(round(.x, 0)))) %>%
  kable(booktabs = T,
        align = "c",
        label = NA,
        caption = "Estimated habitat capacity for the current extent of available habitat to steelhead and potential accessible and suitable habitat if the impoundment were to be removed.") %>%
  kable_styling(full_width = F,
                position = "center",
                bootstrap_options = c("striped", "condensed"))
zena_tbl

```

```{r map, fig.align='center', out.width="100%", fig.cap = "Map of Zena Creek showing the current extent of available habitat to steelhead and potential accessible and suitable habitat if the impoundment were to be removed."}
# simple map
zena_p = zena_qrf_sf %>%
  ggplot() +
  annotation_map_tile(type = "osm", zoomin = 0) +
  geom_sf(aes(color = status), linewidth = 1) +
  # map scale
  annotation_scale(location = "br",
                   width_hint = 0.4) +
  # north arrow
  annotation_north_arrow(location = "br",
                         which_north = "true",
                         pad_x = unit(0, "in"),
                         pad_y = unit(0.35, "in"),
                         style = north_arrow_fancy_orienteering) +
  labs(color = "Status") +
  theme_minimal() +
  theme(axis.text = element_blank())
zena_p

```

### Literature Cited

See, K.E., M.W. Ackerman, R.A. Carmichael, S.L. Hoffmann, and C. Beasley. 2021. Estimating carrying capacity for juvenile salmon using quantile random forest models. Ecosphere 12(3):e03404. 10.1002/ecs2.3404